---
title: "a03 - Practice With Wrangling and Cleaning Data"
subtitle: "Analyzing Salmon Catch Data from Alaska"
author: "Noor Abuzahriyeh"
date: "March 11, 2026"
format:
  html:
    toc: true
    toc-depth: 3
execute:
  warning: false
  message: false
---

# Setup

Load the necessary libraries to load the data

```{r}
library(dplyr) # to reshape data
library(tidyr) # to clean data
library(readr) # to read in data
```

# Introduction

It is important to keep track of which types and how many salmon are caught over the years for ecological and consumer reasons. Firstly, it is important to know if less salmon are being caught, as this can be an indication of if over-fishing has occurred. If the amount of salmon being caught has gradually decreased over the years, the Alaskan fishing authorities may limit the amount of a specific salmon species from being caught, or they may ban salmon catching for a given year altogether. This way, we can prevent any specific species from becoming endangered or going extinct. From a consumer standpoint, we can compare the amount of salmon being caught to the amount sold to understand which species are more popular among the population.

The following analysis that will be done will inform us which species are most abundant in different regions of Alaska by proxy of catch number. The regions included in this study are the Southern Southeast Alaska (SSE), Bristol Bay Management Area (BRB), Northern Southeast Alaska (NSE), Kodiak Management Area (KOD), Prince William Sound Management Area (PWS), and the South Peninsula Subarea (SOP) regions. By checking the salmon catch data from these regions, we would be able to determine which regions have the most salmon or where we are likely to catch more salmon.

# Data

### About the data

The data used in this analysis is the Alaska commercial salmon catches by management region between 1886-1997. It is downloaded from the Knowledge Network for Biocomplexity (KNB), which is an international repository for ecological and environmental data. The data set was published in 2016. I accessed it on March 3, 2026 from this url: <https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1>

#### For my (Noor's) reference: ("Real" analysis with full workflow is under the analysis header)

Load in the data to view it in its original form to be able to understand where the resulting mean_region data below will come from.

```{r}
catch_original <- read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1", method = "libcurl"))

dplyr::glimpse(catch_original)
```

# Analysis

Here, the data is analyzed to determine the average amount of salmon caught across the six regions mentioned previously.

```{r}
# read in the data using read.csv method
# I have a windows CPU, so need to include libcurl method
catch_original <- read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1", method = "libcurl"))

# find the mean catch data by region
mean_region <- catch_original %>% # This is defining a new data frame from the catch_original data frame loaded above.
  select(-All, -notesRegCode) %>%  # this removes the All and notesRegCode from the orginal catch data since All is a summary statistics column, and notesRegCode includes written observations and empty cells.
  mutate(Chinook = if_else(Chinook == "I", "1", Chinook)) %>%  # This will change any letter "I" values to "1" values in the Chinook column.
  mutate(Chinook = as.numeric(Chinook)) %>% # this converts all values in the Chinook column to integers
  pivot_longer(-c(Region,Year), # This changes the data format from wide to long, where the column names become rows and the rows become columns. "Region" and "Year" columns will be excluded.
               names_to = "species", # This creates a new column name called "species" from the columns that are pivoted over.
               values_to = "catch") %>% # This creates a new column called "catch" from the values in the columns that are pivoted over.
  mutate(catch = catch*1000) %>% # This multiplies all the values in the "catch" column by 1000 since the unit of the data is thousands of catches, and it changes the values previously assigned in the catch column to these new values.
  group_by(Region) %>%  # This splits the data to create groups based on the "Region" column.
  summarize(mean_catch = mean(catch)) %>% #  mean(catch) will calculate the mean of the catch data by region since it is grouped together via the pipe, and the results will be added to a column called "mean_catch", as indicated by the assignment operator, "=". Summarize will take those results and create a new table. 
  arrange(desc(mean_catch)) # This will sort the mean_catch data.desc is used to arrange the data in descending order from most on top to least on bottom rather than the default ascending order. 

head(mean_region) # This views the resulting data with the analysis performed above.
```

# Discussion

The mean_region data table above that resulted from the analysis presents the average number of total salmon caught over the years depending on the management region. While it does not tell us which species was the most abundant, it does tell us which regions overall have the highest number of catches. For example, it appears that in the Southern Southeast Alaska (SSE) region, the average number of salmon caught was the highest, at `r mean_region$mean_catch[1]` compared to the lowest number of catches being in the South Peninsula Subarea (SOP) at `r mean_region$mean_catch[6]` , roughly 3 times less than in the SSE region. This difference necessitates the investigation of latitudinal effects on climate, environment, and salmon habits. Based on the 1999 Regional Information Report No. 5J99-05 by Mike Byerly et al. SSE and SOP are at different latitudinal ends in Alaska, which could be an indication that the environment in those waters is less optimal for salmon. Overall, the different regions with Alaska may have more favorable habitat conditions salmon as indicated by the mean amount of salmon that has been caught over the years.
